---
title: "Marine Biofuel ILUC Prototype"
subtitle: "Analysis Report"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
    embed-resources: false
---

## Overview

This report documents the marine biofuel indirect land use change (ILUC) prototype analysis.

## Background

Indirect land use change (ILUC) is a key factor in evaluating the lifecycle greenhouse gas emissions of biofuels. This prototype explores ILUC dynamics in the context of marine biofuel production.

## Methods

*Methods will be described here.*

## Results

### ICCT Table 1: blend ratios and volumes

The chunk below scrapes **Table 1** (page 5) of the ICCT research brief
(*Estimating demand under the IMO Net-Zero Framework*) and reshapes it into a
clean tibble. Each year appears on its own line in the extracted text, so we
pull the numeric tokens per row and split the two compound columns
("% energy demand / billion liters") into separate fields.

```{r}
#| label: scrape-icct-table-1
#| message: false
#| warning: false

# Package versions are pinned via renv (see renv.lock); run renv::restore() to install.
# Functions are called with explicit pkg:: prefixes rather than attaching via library().

pdf_path <- file.path(
  "data",
  "icct_biofuel_demand_report",
  "ID-450-Vegetable-Oil_research-brief_final.pdf"
)

# Extract page 5 as text and keep only the data rows (years 2028–2035)
page5 <- pdftools::pdf_text(pdf_path)[5]
lines <- stringr::str_trim(stringr::str_split(page5, "\n")[[1]])
data_lines <- stringr::str_subset(lines, "^20(2[89]|3[0-5])\\b")

# Each data row yields exactly 10 numeric tokens:
#   year, shipping demand, base target, fossil LNG %,
#   UCO-RD %, UCO-RD bl, VO-RD %, VO-RD bl, UCO demand bl, VO demand bl
nums <- stringr::str_extract_all(data_lines, "[0-9]+\\.?[0-9]*")
mat <- do.call(rbind, lapply(nums, function(x) as.numeric(x[seq_len(10)])))

icct_table_1 <- tibble::as_tibble(mat, .name_repair = "minimal")
names(icct_table_1) <- c(
  "year",
  "shipping_energy_demand_ej",
  "base_target_trajectory_gco2e_mj",
  "fossil_lng_pct",
  "ucord_pct_energy_demand",
  "ucord_billion_liters",
  "vord_pct_energy_demand",
  "vord_billion_liters",
  "uco_demand_billion_liters",
  "vo_demand_billion_liters"
)

icct_table_1 <- icct_table_1 |>
  dplyr::mutate(year = as.integer(year)) |>
  dplyr::arrange(year)

icct_table_1
```

### Interactive table

An interactive, sortable/searchable view of the scraped table.

```{r}
#| label: icct-table-1-interactive
#| message: false
#| warning: false

DT::datatable(
  icct_table_1,
  rownames = FALSE,
  filter = "top",
  options = list(pageLength = 10, scrollX = TRUE)
)
```

### Interactive time series: biofuel demand

Projected biofuel demand by feedstock (used cooking oil vs. vegetable oil),
in billion liters, across the 2028–2035 horizon. Hover for values.

```{r}
#| label: icct-demand-timeseries
#| message: false
#| warning: false

ts_data <- icct_table_1 |>
  dplyr::select(
    year,
    uco_demand_billion_liters,
    vo_demand_billion_liters
  )

dygraphs::dygraph(
  ts_data,
  main = "Projected demand of soy-based biodiesel",
  ylab = "Billion liters"
) |>
  dygraphs::dySeries("uco_demand_billion_liters", label = "UCO demand") |>
  dygraphs::dySeries("vo_demand_billion_liters", label = "VO demand") |>
  dygraphs::dyOptions(strokeWidth = 2, drawPoints = TRUE, pointSize = 3) |>
  dygraphs::dyLegend(show = "always") |>
  dygraphs::dyHighlight(highlightSeriesOpts = list(strokeWidth = 3))
```

## Discussion

*Discussion of findings will go here.*

## References

*References will be listed here.*
