---
title: "Assessing the ILUC implications of increasing marine biofuel demand"
subtitle: "Prototype Proof-of-Concept Analysis"
authors: Environmental Markets Lab (emLab) at University of California, Santa Barbara (UCSB). For International Council on Clean Transportation (ICCT)
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
    embed-resources: false
bibliography: references.bib
---

## Overview

This report documents a prototype proof-of-concept analysis investigating the projected indirect land use change (ILUC) implications of increasing marine biofuel demand. This prototype is being done by Environmental Markets Lab (emLab) at University of California, Santa Barbara (UCSB) for International Council on Clean Transportation (ICCT). All results presented in this analysis should be considered exploratory and preliminary.

## Background

Indirect land use change (ILUC) occurs when demand for crop-based biofuel feedstocks raises agricultural commodity prices, and higher prices push cropland expansion into forest and other carbon-rich land elsewhere in the world. These emissions fall outside the fuel's direct supply chain, so lifecycle accounting that stops at the refinery gate misses them, even though they can rival or exceed the emissions of the fossil fuel being displaced. The current draft of the IMO Net-Zero Framework contains no ILUC safeguards, which makes this question directly relevant to marine fuel policy.

Our analysis traces this chain for marine biofuels:

```{mermaid}
flowchart LR
  A["IMO Net-Zero Framework<br/><i>policy</i>"] --> F["Marine renewable-diesel<br/>(VO-RD) fuel demand<br/><i>ICCT cost-optimization</i>"]
  F --> B["Virgin vegetable-oil<br/>feedstock demand D<br/><i>1.22 kg oil per kg RD</i>"]
  B --> C["Vegetable-oil<br/>price shock %ΔP<br/><i>this report</i>"]
  C --> D["ILUC area and<br/>carbon emissions<br/><i>our land-use model, next</i>"]
```

### Literature review

We draw on four sources, each covering a different link in the chain.

| Source | Role in our pipeline | Key values we use |
|---|---|---|
| ICCT demand brief [@seber2025] | Projects the demand shock $D_t$ | 0 → 139 billion L virgin VO, 2028–2035 |
| ICCT climate & revenue brief [@sturrup2026] | Benchmark fixed-factor translation of $D_t$ into ILUC emissions | 62 g CO₂e/MJ factor; ~950 Mt CO₂e cumulative |
| Chen, Sexton & Smith [@chen2026] | Empirical demand → price → deforestation chain (the design we emulate) | Price multiplier $M \approx 2.6$; ILUC CI 84.1 g CO₂e/MJ |
| Roberts & Schlenker [@roberts2013] | Identification strategy and elasticities for the demand → price step | $\varepsilon_s \approx 0.10$, $\varepsilon_d \approx -0.05$; $M \approx 6.7$ |

: Roles of the four sources in our analysis pipeline.

#### Estimating the maximum potential vegetable oil demand for international shipping under the IMO Net-Zero Framework [@seber2025]

This brief estimates how much vegetable oil demand the Net-Zero Framework could induce. It does not quantify the resulting ILUC emissions, but it flags the risk and provides the demand projection that our analysis and the companion brief below both build on.

- **Policy mechanism.** The framework sets two annual greenhouse-gas fuel-intensity (GFI) targets against a 2008 baseline of 93.3 g CO₂e/MJ: a Base target, with a \$380/t CO₂e Tier 2 penalty for exceeding it, and a stricter Direct Compliance target, with a \$100/t Tier 1 penalty between the two. Fuels are judged only on their direct life-cycle emissions; there are no ILUC safeguards, so crop-based biofuels are an attractive compliance option.
- **Demand derivation.** ICCT builds a least-cost sector-wide fuel mix for 2028–2035. Fossil LNG enters exogenously (0.81 EJ, about 8% of energy demand by 2035, from their POLARIS model), and biofuels are then blended until the fleet-wide average carbon intensity just meets the Base target. Since blending costs less than \$380/t, the cost-minimizing strategy is to meet the Base target and pay only the Tier 1 penalty.
- **Feedstock logic.** The cheapest pathway, used-cooking-oil renewable diesel (UCO-RD, 13.1 g CO₂e/MJ), is blended first up to its supply limit; virgin vegetable-oil renewable diesel (soy-RD, 42.3 g CO₂e/MJ) fills the rest. UCO supply is the binding constraint — roughly 14.3 billion L of feedstock, or 13.6 billion L of fuel, covering only about 5% of shipping energy demand — so soy oil absorbs the remaining compliance burden.
- **Key finding.** VO-RD grows to 46% of shipping energy demand by 2035, implying virgin vegetable-oil demand of up to 139 billion liters (Table 1, p. 5), roughly 3.5 times the 40 billion liters of vegetable oil used for biofuels today. ICCT judges existing virgin oil supply insufficient to meet the targets beyond 2032 without new cropland or diversion from other sectors.
- **Use in this analysis.** We scrape Table 1 and use the `vo_demand_billion_liters` column as the demand shock for our land-use model (see [Data sources](#data-sources)).

#### Climate and revenue considerations for the IMO Net-Zero Framework [@sturrup2026]

This companion brief takes the demand projection from @seber2025 and translates it into ILUC emissions — the same step our prototype performs, using different methods. It covers two topics: biofuel use for compliance, and the climate finance raised from penalties.

- **Approach.** They assume the projected vegetable-oil demand is met with soy-RD and apply a single illustrative ILUC emission factor of 62 g CO₂e/MJ from the U.S. EPA's Model Comparison Exercise [@epa2023mce]. Since the blend is assumed to meet the Base target, the ILUC-adjusted carbon intensity of the fuel mix is simply the Base-target GFI plus the ILUC contribution.
- **Provenance of the 62.** The value is the GCAM estimate of land-use-change carbon intensity for soybean oil biodiesel in Table 7.7-1 of the EPA exercise; the other models report 295 (ADAGE), 10 (GTAP), 23 (GLOBIOM), and 10 (GREET). The exercise ran a 1 billion gallon per year soybean-biodiesel shock through five models, and the spread is a fair measure of how unsettled ILUC estimates remain. One unit caveat: the EPA table is in kg CO₂e/MMBTU (62 kg CO₂e/MMBTU ≈ 59 g CO₂e/MJ), while the ICCT brief cites the figure as 62 g CO₂e/MJ.
- **Caveat on the factor.** The value may be high for U.S. soy but likely understates ILUC for Indonesian palm or Brazilian soy; GLOBIOM-based estimates for the EU exceed 100 g CO₂e/MJ for high-risk feedstocks [@valin2015]. It is an illustrative point estimate, not a bound.
- **ILUC findings.** The ILUC contribution to fleet GFI grows from 0.9 g CO₂e/MJ in 2029 to 28.5 by 2035 (their Table A1), pushing the ILUC-adjusted GFI to 93.9 g CO₂e/MJ — above the 93.3 fossil baseline. A compliant fuel mix could be worse for the climate than the fuel it replaces.
- **Magnitude.** Annual ILUC emissions reach roughly 285 Mt CO₂e by 2035; the 2028–2035 cumulative total of about 950 Mt CO₂e is comparable to a full year of well-to-wake emissions from international shipping (939 Mt CO₂e in 2024).
- **Revenue interaction.** Blending biofuels to avoid Tier 2 penalties erodes Net-Zero Fund revenue (projected at \$11–12 billion per year), and raising the Tier 1 price would make blending more attractive still, cutting revenue further while worsening ILUC.
- **Policy conclusion.** They recommend ILUC accounting in the IMO LCA guidelines (as in CORSIA and California's LCFS), limits on food- and feed-based biofuels, and financial incentives reserved for advanced fuels.

#### Using vegetable oils for biofuel accelerates tropical deforestation and increases carbon emissions [@chen2026]

This working paper is the closest empirical analog to our prototype. Rather than assuming an ILUC factor, the authors estimate the full demand → price → land conversion → emissions chain econometrically; our analysis aims to reproduce that structure for the marine case.

- **Method.** Three steps: demand and supply models for the major oil crops (palm, soybean, rapeseed, sunflower) establish how biomass-based diesel (BBD) demand moves palm prices; a panel regression on 32,394 grid cells (9 km × 9 km, 2002–2018) of remote-sensing data estimates the effect of palm prices on deforestation, with fixed effects plus reverse-causality and weather checks; and spatial biomass, peatland, and forest-type data convert deforested area into carbon emissions.
- **Price response.** BBD averaged 8.5% of global vegetable-oil consumption over 2002–2018, which the authors estimate raised palm oil prices 21.2% over business-as-usual (90% CI: 12.2–28.5%). The key parameter is a demand flexibility of 3.5: a 1% drop in oil-crop production raises palm prices 3.5%.
- **Supply response.** Short-run palm supply elasticity is small and insignificant (palm is a perennial that takes about four years to fruit); the long-run elasticity is about 0.1. Some 84% of the competing-crop supply response comes from area expansion and only 16% from yield, at odds with the strong intensification response many CGE-based ILUC models assume.
- **Deforestation.** A \$1,000/ha increase in attainable palm revenue drives about 9.3 additional hectares of deforestation per average grid cell, which aggregates to 1.7 million ha of forest-to-palm conversion (90% CI: 1.0–2.3 Mha) — about 20% of all such conversion in Indonesia and Malaysia over the period. Conversion falls mostly on natural forest and is driven mainly by industrial growers.
- **Emissions.** Net land-use-change emissions are 1,086 Tg CO₂e (biomass loss 758, peat 614, soil carbon 95, less 380 sequestered by the palms), amortized over 25 years. That works out to an ILUC carbon intensity of 84.1 g CO₂e/MJ (90% CI: 48.7–113.5); with cultivation, processing, and transport added (34.6–63.5), total carbon intensity is 118.7–147.6 g CO₂e/MJ, or 22–52% above petroleum diesel (~97). The authors treat this as a conservative lower bound, since it excludes ILUC from other oils and from palm outside the study region.
- **Substitution.** Vegetable oils are close substitutes, so policies that restrict only palm still bid up palm prices by diverting demand to soy, rapeseed, and sunflower. This is the same market linkage behind the soy-driven ILUC concern in the ICCT briefs, and it is why the authors favor waste feedstocks (UCO, tallow) that avoid land-use effects.

#### Identifying supply and demand elasticities of agricultural commodities: implications for the US ethanol mandate [@roberts2013]

This paper supplies the identification strategy and elasticity estimates for the demand-to-price step: how a mandated biofuel demand shock translates into an equilibrium commodity price change.

- **Identification.** The paper uses competitive storage theory to justify weather-driven yield shocks as instruments: past shocks shift inventories and futures prices, identifying supply, while contemporaneous shocks move current prices, identifying demand. First-stage F-statistics exceed 10. The analysis covers corn, wheat, rice, and soybeans (about 75% of global food calories), aggregated into caloric equivalents.
- **Elasticities.** Supply is 0.09–0.12 and demand −0.03 to −0.07 (their Table 1), both larger in magnitude than uninstrumented estimates. The implied price multiplier, 1/(β~s~ − β~d~), is 5.8–7.7: small demand shifts produce large price changes.
- **Extensive margin.** The land-area elasticity is 0.07–0.08, with no significant fertilizer response. Added production comes from expanding cropland rather than intensification, which is the mechanism connecting biofuel demand to land-use change.
- **Ethanol-mandate application.** Diverting about 5% of world caloric production (roughly one-third of US corn) to ethanol raises staple food prices about 30% (95% CI: 14–35%; closer to 20% if a third of the calories return as feed byproduct), expands world crop area about 2%, and implies a consumer-surplus loss near \$180 billion per year, borne mostly by net-food-importing developing countries.
- **Use in this analysis.** We take the elasticity estimates and the multiplier logic from this paper to map the marine demand shock into a vegetable-oil price change; land conversion then follows through the mechanism in @chen2026.

## Methods

### From demand projection to a vegetable-oil price shock

Our land-use model takes as input a price shock: the proportional change in the vegetable-oil price attributable to new marine biofuel demand. We derive it from the ICCT demand series (`vo_demand_billion_liters`; see [Data sources](#data-sources)) using the standard comparative-statics result that an outward demand shift moves the equilibrium price by the size of the shift divided by the gap between the supply and demand elasticities [@roberts2013]:

$$ \%\Delta P_t \;\approx\; \underbrace{\frac{D_t}{Q_0}}_{\text{demand shift}} \;\times\; \underbrace{\frac{1}{\varepsilon_s - \varepsilon_d}}_{\text{price multiplier } M} $$

where $D_t$ is the projected marine virgin vegetable-oil demand in year $t$ (billion L), $Q_0$ is baseline global vegetable-oil consumption (billion L), $\varepsilon_s > 0$ is the supply elasticity, and $\varepsilon_d < 0$ is the demand elasticity (so the denominator is $\varepsilon_s + |\varepsilon_d|$). The shock depends on demand as a share of the whole market, not on absolute volume.

We take elasticities from the two empirical papers, which bracket the plausible range:

- **Roberts & Schlenker [@roberts2013]** — inelastic staple-crop demand ($\varepsilon_d \approx -0.05$) with $\varepsilon_s \approx 0.10$, giving a large multiplier $M \approx 6.7$.
- **Chen, Sexton & Smith [@chen2026]** — more elastic vegetable-oil demand ($\varepsilon_d \approx -0.29$, from their demand flexibility of ~3.5) with $\varepsilon_s \approx 0.10$, giving $M \approx 2.6$.

The difference between the two comes almost entirely from the demand side: staple food demand is much less elastic than vegetable-oil demand, since the oils trade as a fungible complex. As a check on the formula and our parameter choices, it recovers both papers' headline results: 8.5% × 2.6 ≈ 22% against Chen's reported 21.2%, and 5% × 6.7 ≈ 33% against Roberts & Schlenker's reported 29–39%. We repeat these checks in code with the results.

A few modeling choices matter for interpretation:

- **Market definition.** Because vegetable oils are close substitutes, the shock plays out in the aggregate oil complex, with palm as the marginal price, even though the marine demand lands on soy [@chen2026]. We therefore set $Q_0$ to total global vegetable-oil consumption rather than the volume currently used for biofuels.
- **Large shocks.** The constant-elasticity approximation is a marginal result. From about 2032 the ICCT demand becomes a large fraction of the market, where the linear form loses reliability and the true price response is likely convex, consistent with the positively skewed confidence interval in @roberts2013 and with the supply shortfall ICCT itself flags.
- **Short- vs. long-run.** Oil palm is a perennial with roughly a four-year lag to fruit, so short-run supply is nearly fixed and the long-run response is larger [@chen2026]. The elasticities above are long-run values.
- **Soy co-product.** Crushing soybeans for oil also produces protein meal, and meal revenue partially offsets the oil-price pressure — the analogue of the distillers-grains adjustment in @roberts2013, which scales their price effect by roughly two-thirds under full feed recycling. Chen et al. avoid the issue by pricing the shock off palm. We apply no adjustment, so on this margin our estimates err high.

Converting $\%\Delta P_t$ into ILUC area and emissions is the job of our own land-use regression, which we will develop next. The price-shock series is written to `data/derived/vegetable_oil_price_shock.csv` for that step.

## Data sources

### Projected demand for soy-based vegetable oil (from ICCT)

In the research brief
[Estimating the maximum potential vegetable oil demand for international shipping under the International Maritime Organization’s Net-Zero Framework](https://theicct.org/wp-content/uploads/2025/09/ID-450-Vegetable-Oil_research-brief_final.pdf) [@seber2025], ICCT estimates the projected demand for marine renewable biodiesel (and the associated demand for used cooking oil feedstock and virgin soy-based vegetable oil feedstock) under the proposed IMO Net-Zero Framework, from 2028 to 2035. This projected demand is summarized in the report's **Table 1** (page 5). We scraped the PDF report and turned this into a data tibble for use in our analysis:
```{r}
#| label: scrape-icct-table-1
#| message: false
#| warning: false

# Package versions are pinned via renv (see renv.lock); run renv::restore() to install.
# Functions are called with explicit pkg:: prefixes rather than attaching via library().

pdf_path <- file.path(
  "data",
  "icct_biofuel_demand_report",
  "ID-450-Vegetable-Oil_research-brief_final.pdf"
)

# Extract page 5 as text and keep only the data rows (years 2028–2035)
page5 <- pdftools::pdf_text(pdf_path)[5]
lines <- stringr::str_trim(stringr::str_split(page5, "\n")[[1]])
data_lines <- stringr::str_subset(lines, "^20(2[89]|3[0-5])\\b")

# Each data row yields exactly 10 numeric tokens:
#   year, shipping demand, base target, fossil LNG %,
#   UCO-RD %, UCO-RD bl, VO-RD %, VO-RD bl, UCO demand bl, VO demand bl
# Each year appears on its own line in the extracted text, so we
#pull the numeric tokens per row and split the two compound columns
# ("% energy demand / billion liters") into separate fields.
nums <- stringr::str_extract_all(data_lines, "[0-9]+\\.?[0-9]*")
mat <- do.call(rbind, lapply(nums, function(x) as.numeric(x[seq_len(10)])))

icct_table_1 <- tibble::as_tibble(mat, .name_repair = "minimal")
names(icct_table_1) <- c(
  "year",
  "shipping_energy_demand_ej",
  "base_target_trajectory_gco2e_mj",
  "fossil_lng_pct",
  "ucord_pct_energy_demand",
  "ucord_billion_liters",
  "vord_pct_energy_demand",
  "vord_billion_liters",
  "uco_demand_billion_liters",
  "vo_demand_billion_liters"
)

icct_table_1 <- icct_table_1 |>
  dplyr::mutate(year = as.integer(year)) |>
  dplyr::arrange(year)
```

#### Column schema

The scraped columns, their meaning, and units (derived from Table 1 and its
notes in the ICCT brief [@seber2025]). "UCO-RD" is used-cooking-oil-based
renewable diesel and "VO-RD" is vegetable-oil-based renewable diesel; the
*-RD* volumes are the fuel produced, while the *UCO/VO demand* columns are the
feedstock required to make it.

```{r}
#| label: icct-table-1-schema
#| message: false
#| warning: false

icct_table_1_schema <- tibble::tribble(
  ~column                           , ~description                                                                                     , ~units               ,
  "year"                            , "Calendar year of the projection"                                                                , "year (2028–2035)"   ,
  "shipping_energy_demand_ej"       , "Total international shipping energy demand (ships ≥5,000 GT; ICCT POLARIS model)"               , "EJ (exajoules)"     ,
  "base_target_trajectory_gco2e_mj" , "Base target GHG fuel-intensity trajectory the blended fuel must meet in the Net-Zero Framework" , "g CO2e/MJ"          ,
  "fossil_lng_pct"                  , "Fossil LNG share of shipping energy demand"                                                     , "% of energy demand" ,
  "ucord_pct_energy_demand"         , "UCO-RD share of shipping energy demand"                                                         , "% of energy demand" ,
  "ucord_billion_liters"            , "UCO-RD fuel volume blended into the fuel mix"                                                   , "billion liters"     ,
  "vord_pct_energy_demand"          , "VO-RD share of shipping energy demand"                                                          , "% of energy demand" ,
  "vord_billion_liters"             , "VO-RD fuel volume blended into the fuel mix"                                                    , "billion liters"     ,
  "uco_demand_billion_liters"       , "Used cooking oil (UCO) feedstock demand to produce the UCO-RD"                                  , "billion liters"     ,
  "vo_demand_billion_liters"        , "Virgin vegetable oil (VO) feedstock demand to produce the VO-RD"                                , "billion liters"
)

knitr::kable(
  icct_table_1_schema,
  col.names = c("Column", "Description", "Units")
)
```

```{r}
#| label: icct-table-1-interactive
#| message: false
#| warning: false

DT::datatable(
  icct_table_1,
  rownames = FALSE,
  filter = "top",
  caption = 'Caption from the ICCT report: "UCO-RD and VO-RD blend ratios and volumes to comply with the base target emissions trajectory. Notes: UCO-based renewable diesel availability is used as a constraint, and the rest of the demand is assumed to come from soy oil-based renewable diesel. UCO and VO demand to reach the estimated renewable diesel volumes are included. Estimations for shipping energy demand for ships of ≥ 5,000 gross tonnage are from the ICCT’s POLARIS Mode"',
  options = list(pageLength = 10, scrollX = TRUE)
)
```

The most important metric for our analysis is `vo_demand_billion_liters` (the projected soy-based virgin vegetable oil feedstock demand). It is this projected demand that we believe will most directly lead to a price shock that can be used in our land use model. Here we look at a time series of this projected demand:

```{r}
#| label: icct-demand-timeseries
#| message: false
#| warning: false

ts_data <- icct_table_1 |>
  dplyr::select(
    year,
    vo_demand_billion_liters
  )

dygraphs::dygraph(
  ts_data,
  main = "Projected demand for soy-based vegetable oil virgin feedstock",
  ylab = "Billion liters"
) |>
  dygraphs::dySeries("vo_demand_billion_liters", label = "VO demand") |>
  dygraphs::dyOptions(strokeWidth = 2, drawPoints = TRUE, pointSize = 3) |>
  dygraphs::dyLegend(show = "always") |>
  dygraphs::dyHighlight(highlightSeriesOpts = list(strokeWidth = 3))
```

## Results

### Derived price shock

We now apply the mapping from [Methods](#methods) to the ICCT demand series to estimate the annual vegetable-oil price shock attributable to marine biofuel demand. The only assumptions are the baseline market size $Q_0$ and the elasticity scenarios, set explicitly below.

```{r}
#| label: price-shock-parameters
#| message: false
#| warning: false

# --- Parameters (see Methods for sourcing and caveats) ---

# Baseline global vegetable-oil consumption, billion liters.
# ~225 Mt (USDA, major vegetable oils, 2023/24) at ~0.92 kg/L.
q0_billion_liters <- 245

# Elasticity scenarios. eps_d is negative, so the multiplier denominator
# is eps_s + |eps_d|. The scenarios bracket the plausible demand response:
# vegetable oil trades as a fungible complex (more elastic demand, smaller
# price response) while staple food demand is very inelastic (larger one).
eps_s_chen <- 0.10 # long-run oil-crop supply elasticity (Chen et al. 2026)
eps_d_chen <- -0.29 # vegetable-oil demand elasticity (1 / 3.5 flexibility)
eps_s_rs <- 0.10 # staple-crop supply elasticity (Roberts & Schlenker 2013)
eps_d_rs <- -0.05 # staple-crop demand elasticity

m_chen <- 1 / (eps_s_chen - eps_d_chen) # ~2.6
m_rs <- 1 / (eps_s_rs - eps_d_rs) # ~6.7

scenario_params <- tibble::tribble(
  ~scenario, ~source,                                              ~eps_s,     ~eps_d,     ~multiplier,
  "Low",     "Chen, Sexton & Smith (2026): vegetable-oil complex", eps_s_chen, eps_d_chen, m_chen,
  "High",    "Roberts & Schlenker (2013): staple-food calories",   eps_s_rs,   eps_d_rs,   m_rs
)

knitr::kable(
  scenario_params,
  digits = 2,
  col.names = c(
    "Scenario", "Elasticity source",
    "Supply elasticity (ε_s)", "Demand elasticity (ε_d)", "Multiplier M"
  )
)
```

Before applying the multipliers to our series, we check that each one recovers its source paper's own headline price effect:

```{r}
#| label: price-shock-validation
#| message: false
#| warning: false

validation <- tibble::tribble(
  ~check,                                                        ~shock_pct, ~m,     ~reported,
  "Chen et al.: BBD was 8.5% of vegetable-oil use, 2002-2018",   8.5,        m_chen, "21.2% (90% CI: 12.2-28.5%)",
  "Roberts & Schlenker: ethanol mandate = 5% of world calories", 5.0,        m_rs,   "~30% (range 29-39%)"
) |>
  dplyr::mutate(implied_pct = round(shock_pct * m, 1)) |>
  dplyr::select(check, shock_pct, implied_pct, reported)

knitr::kable(
  validation,
  col.names = c(
    "Validation case", "Demand shock (% of market)",
    "Implied %ΔP (our formula)", "Paper's reported %ΔP"
  )
)
```

Both checks land within the papers' reported ranges. Applying the multipliers to the ICCT series:

```{r}
#| label: derive-price-shock
#| message: false
#| warning: false

# %ΔP_t = (D_t / Q0) * M, expressed as a percent. The demand share is the
# outward demand shift as a fraction of the whole vegetable-oil market.
# beyond_marginal flags years where the shift exceeds ~25% of the market,
# outside the regime where a constant-elasticity approximation is reliable.
price_shock <- icct_table_1 |>
  dplyr::transmute(
    year,
    vo_demand_billion_liters,
    demand_share_pct = 100 * vo_demand_billion_liters / q0_billion_liters,
    pct_delta_p_low = demand_share_pct * m_chen,
    pct_delta_p_high = demand_share_pct * m_rs,
    beyond_marginal = demand_share_pct > 25
  )
```

```{r}
#| label: price-shock-table
#| message: false
#| warning: false

DT::datatable(
  price_shock |>
    dplyr::mutate(
      dplyr::across(where(is.numeric), \(x) round(x, 1)),
      beyond_marginal = ifelse(beyond_marginal, "yes", "")
    ),
  rownames = FALSE,
  colnames = c(
    "Year",
    "VO demand (billion L)",
    "Demand share of market (%)",
    "%ΔP low (Chen, M≈2.6)",
    "%ΔP high (Roberts-Schlenker, M≈6.7)",
    "Beyond marginal regime"
  ),
  caption = "Projected vegetable-oil price shock attributable to marine biofuel demand. The low/high columns bracket the price response using the two elasticity scenarios. Years flagged 'beyond marginal regime' (demand share > 25% of the market, from ~2032) exceed the range where the constant-elasticity approximation is reliable; treat those values as indicative of direction and scale, not precise magnitudes.",
  options = list(pageLength = 10, scrollX = TRUE, dom = "t")
)
```

```{r}
#| label: price-shock-timeseries
#| message: false
#| warning: false

dygraphs::dygraph(
  price_shock |> dplyr::select(year, pct_delta_p_low, pct_delta_p_high),
  main = "Projected vegetable-oil price shock from marine biofuel demand",
  ylab = "% change in vegetable oil price"
) |>
  dygraphs::dySeries(
    "pct_delta_p_high",
    label = "High (Roberts-Schlenker, M≈6.7)"
  ) |>
  dygraphs::dySeries("pct_delta_p_low", label = "Low (Chen, M≈2.6)") |>
  dygraphs::dyShading(from = 2031.5, to = 2035.5) |>
  dygraphs::dyOptions(strokeWidth = 2, drawPoints = TRUE, pointSize = 3) |>
  dygraphs::dyLegend(show = "always") |>
  dygraphs::dyHighlight(highlightSeriesOpts = list(strokeWidth = 3))
```

The shaded region (2032 on) marks years where marine demand exceeds a quarter of the baseline market. The linear approximation is unreliable there and, if anything, understates the true response. Last, we write the series out for the land-use step:

```{r}
#| label: export-price-shock
#| message: false
#| warning: false

# Hand-off to the price -> ILUC regression (full precision, no rounding).
dir.create(file.path("data", "derived"), showWarnings = FALSE, recursive = TRUE)
utils::write.csv(
  price_shock,
  file.path("data", "derived", "vegetable_oil_price_shock.csv"),
  row.names = FALSE
)
```

The exported columns `pct_delta_p_low` and `pct_delta_p_high` (in `data/derived/vegetable_oil_price_shock.csv`) are the inputs to the next step of the analysis: converting the price shock into ILUC area and emissions.

## Discussion

*Discussion of findings will go here.*

## References

::: {#refs}
:::
