---
title: "Assessing the ILUC implications of increasing marine biofuel demand"
subtitle: "Prototype Proof-of-Concept Analysis"
authors: Environmental Markets Lab (emLab) at University of California, Santa Barbara (UCSB). For International Council on Clean Transportation (ICCT)
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
    embed-resources: false
bibliography: references.bib
---

## Overview

This report documents a prototype proof-of-concept analysis investigating the projected indirect land use change (ILUC) implications of increasing marine biofuel demand. This prototype is being done by Environmental Markets Lab (emLab) at University of California, Santa Barbara (UCSB) for International Council on Clean Transportation (ICCT). All results presented in this analysis should be considered exploratory and preliminary.

## Background

Indirect land use change (ILUC) occurs when policy-induced demand for crop-based biofuel feedstocks raises agricultural commodity prices, and those higher prices induce the conversion of forest and other carbon-rich land to cropland elsewhere in the world. Because the carbon released by that conversion sits outside any fuel's direct supply chain, it is invisible to lifecycle accounting that stops at the refinery gate — yet it can rival or exceed the emissions of the fossil fuel being displaced. ILUC is therefore a key factor in evaluating the lifecycle greenhouse gas emissions of biofuels, and a central open question for the IMO Net-Zero Framework, whose current draft contains no ILUC safeguards.

This prototype traces that causal chain for marine biofuels:

```{mermaid}
flowchart LR
  A["IMO Net-Zero Framework<br/><i>policy</i>"] --> B["Marine vegetable-oil<br/>feedstock demand D<br/><i>ICCT projection</i>"]
  B --> C["Vegetable-oil<br/>price shock %ΔP<br/><i>this report</i>"]
  C --> D["ILUC area and<br/>carbon emissions<br/><i>our land-use model, next</i>"]
```

### Literature review

Four sources anchor the analysis, each supplying a different link in the chain above.

| Source | Role in our pipeline | Key values we use |
|---|---|---|
| ICCT demand brief [@seber2025] | Projects the demand shock $D_t$ | 0 → 139 billion L virgin VO, 2028–2035 |
| ICCT climate & revenue brief [@sturrup2026] | Benchmark fixed-factor translation of $D_t$ into ILUC emissions | 62 g CO₂e/MJ factor; ~950 Mt CO₂e cumulative |
| Chen, Sexton & Smith [@chen2026] | Empirical demand → price → deforestation chain (the design we emulate) | Price multiplier $M \approx 2.6$; ILUC CI 84.1 g CO₂e/MJ |
| Roberts & Schlenker [@roberts2013] | Identification strategy and elasticities for the demand → price step | $\varepsilon_s \approx 0.10$, $\varepsilon_d \approx -0.05$; $M \approx 6.7$ |

: Roles of the four sources in our analysis pipeline.

#### Estimating the maximum potential vegetable oil demand for international shipping under the IMO Net-Zero Framework [@seber2025]

This brief estimates the *demand* for vegetable oil that the Net-Zero Framework could induce; it stops short of quantifying the resulting ILUC emissions, but it flags the ILUC risk and supplies the demand projection that our analysis (and the companion ICCT brief below) build on.

- **Policy mechanism.** The Net-Zero Framework sets two annual greenhouse-gas fuel-intensity (GFI) targets against a 2008 baseline of 93.3 g CO₂e/MJ: a Base target (Tier 2 penalty of \$380/t CO₂e for exceeding it) and a stricter Direct Compliance target (Tier 1 penalty of \$100/t CO₂e between the two). The framework is technology-neutral and, as currently drafted, judges fuels only on direct life-cycle emissions — with **no ILUC safeguards** — creating the incentive to blend crop-based biofuels.
- **How demand is derived (cost-optimization).** They build a least-cost sector-wide fuel mix from 2028 through 2035 in two steps: (1) fossil LNG is added exogenously (~0.81 EJ, ~8.1% of energy demand by 2035, from the POLARIS model); (2) biofuels are blended so the fleet-wide weighted-average carbon intensity just meets the Base target GFI, avoiding the \$380/t Tier 2 penalty. Because blending cost stays below \$380/t, the optimal strategy is to meet the Base target and pay only the Tier 1 penalty.
- **Feedstock logic.** The cheapest pathway — used-cooking-oil renewable diesel (UCO-RD, CI ≈ 13.1 g CO₂e/MJ) — is blended first up to its supply limit, then virgin vegetable-oil RD (soy-RD, CI ≈ 42.3 g CO₂e/MJ) fills the rest. UCO supply is the binding constraint (~14.3 billion L, ~13.6 billion L of RD, enough for only ~4.9% of shipping energy demand), so soy oil absorbs the remaining compliance burden.
- **Key finding.** VO-RD rises to ~46% of shipping energy demand by 2035, implying virgin vegetable-oil demand of up to **~139 billion liters by 2035** (**Table 1**, p. 5) — about 3.5× the ~40 billion L of vegetable oil used in biofuels today. Existing virgin VO supply is judged insufficient to meet targets beyond 2032 without new cropland or diversion from other sectors.
- **Use in this analysis.** We scrape Table 1's `vo_demand_billion_liters` and use it as the demand shock feeding our land-use model (see [Data sources](#data-sources)).

#### Climate and revenue considerations for the IMO Net-Zero Framework [@sturrup2026]

Companion ICCT brief that **takes the [@seber2025] demand projection and translates it into ILUC emissions** — the step most directly analogous to what our prototype analysis does. It examines two elements of the framework: biofuel use for compliance, and the climate finance raised from penalties.

- **How demand becomes ILUC (their approach).** They take the Seber & Pavlenko [@seber2025] vegetable-oil demand path from Table 1, assume it is met with soy-RD, and apply a single illustrative **ILUC emission factor of 62 g CO₂e/MJ** drawn from the U.S. EPA's Model Comparison Exercise [@epa2023mce]. Because the blend is assumed to meet the Base target, they compute the true fuel carbon intensity by *adding* the soy-RD ILUC emissions on top of the Base-target GFI (i.e., ILUC-adjusted GFI = Base target GFI + ILUC contribution).
- **Tracing the 62.** The value is the **land-use-change carbon intensity of soybean oil biodiesel from the GCAM model** in Table 7.7-1 of the EPA Model Comparison Exercise [@epa2023mce], where GCAM reports 62 (vs. ADAGE 295, GTAP 10, GLOBIOM 23, GREET 10). The exercise shocked soybean-biodiesel demand by 1 billion gallons/yr and ran it through five models; the wide spread reflects genuine ILUC uncertainty. Note a unit caveat: the EPA table reports 62 in **kg CO₂e/MMBTU** (≈ 59 g CO₂e/MJ), whereas the ICCT brief carries the bare figure across as 62 g CO₂e/MJ.
- **Caveat on the factor.** 62 g CO₂e/MJ may be high for U.S. soy but could *understate* ILUC for Indonesian palm or Brazilian soy, where the GLOBIOM analysis for the EU finds ILUC emissions exceeding 100 g CO₂e/MJ for palm and other high-risk feedstocks [@valin2015] — so their estimate is a single illustrative point, not a bound.
- **ILUC findings.** The ILUC contribution to fleet GFI grows from ~0.9 g CO₂e/MJ in 2029 to ~28.5 g CO₂e/MJ by 2035 (Appendix Table A1). This pushes the ILUC-adjusted GFI to **~93.86 g CO₂e/MJ by 2035 — above the 93.3 g CO₂e/MJ 2008 fossil baseline**, meaning the compliant fuel mix could be *worse* for the climate than the fuel it replaces.
- **Magnitude.** Annual ILUC emissions reach ~285 Mt CO₂e by 2035, with cumulative 2028–2035 ILUC emissions of ~950 Mt CO₂e — comparable to the entire ~939 Mt CO₂e annual well-to-wake GHG emissions of international shipping in 2024.
- **Revenue interaction.** Blending biofuels to avoid Tier 2 penalties erodes Net-Zero Fund revenue (projected ~\$11–12 billion/yr); raising the Tier 1 RU price would make biofuel blending *more* attractive, further collapsing revenue while worsening ILUC.
- **Policy conclusion.** They argue for explicit ILUC accounting in the IMO LCA guidelines (as in CORSIA and California's LCFS), limits on food/feed biofuels, and steering incentives toward advanced fuels.

#### Using vegetable oils for biofuel accelerates tropical deforestation and increases carbon emissions [@chen2026]

The closest empirical analog to what this prototype attempts: rather than assuming an ILUC factor, Chen, Sexton & Smith estimate the full **demand → price → land conversion → emissions** chain econometrically. Their pipeline is essentially the causal structure our analysis aims to reproduce for the marine case.

- **Method (a chain of econometric steps).** (1) Estimate demand and supply models for the major oil crops (oil palm, soybean, rapeseed/canola, sunflower) to get how biomass-based diesel (BBD) demand moves palm prices; (2) estimate the effect of palm prices on deforestation using high-resolution remote sensing — **32,394 grid cells (9 km × 9 km), 2002–2018**, with panel fixed effects and reverse-causality/weather robustness checks; (3) combine with spatial biomass, peatland, and forest-type data to convert deforested area into carbon emissions.
- **Price response.** BBD averaged **8.5% of global vegetable-oil consumption** over 2002–2018; the authors estimate this demand raised palm oil prices by **~21.2%** vs. business-as-usual (90% CI: 12.2–28.5%). A key input is the estimated demand flexibility: a 1% drop in oil-crop production raises palm prices ~3.5%.
- **Supply is inelastic and expands via *area*, not yield.** Short-run palm supply elasticity is small/insignificant (palm is perennial, ~4 yrs to fruit); long-run only ~0.1%. Critically, they find **~84% of the competing-crop supply response comes from area expansion and only ~16% from yield/intensification** — directly contradicting the strong intensification response many CGE/ILUC models assume.
- **Deforestation.** A \$1,000/ha rise in attainable palm revenue drives ~9.3 additional ha of deforestation per average cell → **~1.7 million ha of forest-to-palm conversion** (90% CI: 1.0–2.3 Mha), about **20%** of all forest-to-palm conversion in Indonesia and Malaysia over the period. Conversion falls overwhelmingly on natural (unmanaged) forest and is driven mainly by industrial growers.
- **Emissions / carbon intensity.** Net land-use-change emissions ≈ **1,086 Tg CO₂e** (biomass 757.7 + peat 614.0 + non-peat SOC 94.8 − palm sequestration 380.2), amortized over 25 yrs. This is an **ILUC carbon intensity of 84.1 g CO₂e/MJ** (90% CI: 48.7–113.5); adding cultivation/processing/transport (34.6–63.5) gives a total CI of **118.7–147.6 g CO₂e/MJ — 22–52% *above* petroleum diesel** (~97 g CO₂e/MJ). Because it excludes ILUC from other oils and out-of-region palm, they call it a conservative lower bound.
- **Substitution → why targeting palm alone fails.** Vegetable oils are close substitutes, so biofuel policies that restrict only palm still bid up palm prices indirectly by diverting demand to soy/rapeseed/sunflower. This is exactly the market-linkage mechanism behind the ICCT briefs' soy-driven ILUC concern, and it argues for favoring waste feedstocks (UCO, tallow) that avoid adverse land-use change.

#### Identifying supply and demand elasticities of agricultural commodities: implications for the US ethanol mandate [@roberts2013]

Provides the identification strategy and elasticity magnitudes for the price-response step of the ILUC chain — how a mandated biofuel demand shock translates into an equilibrium commodity price change.

- **Identification.** Exploits **competitive storage theory** to use weather-driven yield shocks as instruments: past yield shocks shift inventories and futures prices (identifying *supply*), while contemporaneous shocks move current prices (identifying *demand*). First-stage F-statistics > 10 (strong instruments). Applied to the four staple crops — corn, wheat, rice, soybeans (~75% of global food calories) — aggregated into caloric equivalents.
- **Elasticities (both inelastic; supply > demand).** Supply elasticity **≈ 0.09–0.12** and demand elasticity **≈ −0.03 to −0.07** (Table 1), both larger in magnitude than traditional uninstrumented estimates. The implied **price multiplier 1/(β_s − β_d) ≈ 5.8–7.7**, so small demand shifts produce large price swings.
- **Supply response is on the extensive margin.** Estimated land-**area** elasticity ≈ **0.07–0.08** with no significant fertilizer/intensive-margin response — i.e., added production comes from expanding cropland, the very mechanism that makes biofuel demand drive land-use change.
- **Ethanol-mandate application.** Diverting ~5% of world caloric production (≈ one-third of US corn) to ethanol raises staple food prices by **~30%** (95% CI: 14–35%; ~20% if one-third of calories return as feed byproduct), expands world crop area ~2%, and implies a ~\$180 billion/yr consumer-surplus loss borne disproportionately by net-food-importing developing countries.
- **Use in this analysis.** These are the canonical inelastic-supply/demand estimates and the weather-IV design we draw on to justify how a marine biofuel demand shock maps to a vegetable-oil price change — the input that then drives land conversion in the [@chen2026] mechanism.

## Methods

### From demand projection to a vegetable-oil price shock

Our land-use model requires a **price shock** — the proportional change in the vegetable-oil price attributable to new marine biofuel demand — as its input. We derive that shock from the ICCT projected demand series (`vo_demand_billion_liters`, see [Data sources](#data-sources)) using the standard comparative-statics result that an outward shift in demand moves the equilibrium price by the size of the shift divided by the gap between the supply and demand elasticities [@roberts2013]:

$$ \%\Delta P_t \;\approx\; \underbrace{\frac{D_t}{Q_0}}_{\text{demand shift}} \;\times\; \underbrace{\frac{1}{\varepsilon_s - \varepsilon_d}}_{\text{price multiplier } M} $$

where $D_t$ is the projected marine virgin vegetable-oil demand in year $t$ (billion L), $Q_0$ is baseline global vegetable-oil consumption (billion L), $\varepsilon_s > 0$ is the supply elasticity, and $\varepsilon_d < 0$ is the demand elasticity (so the denominator is $\varepsilon_s + |\varepsilon_d|$). The shock scales with demand *as a share of the whole market*, not with absolute volume.

We source the elasticities from the two empirical references, which bracket the plausible range:

- **Roberts & Schlenker [@roberts2013]** — inelastic staple-crop demand ($\varepsilon_d \approx -0.05$) with $\varepsilon_s \approx 0.10$, giving a large multiplier $M \approx 6.7$.
- **Chen, Sexton & Smith [@chen2026]** — more elastic vegetable-oil demand ($\varepsilon_d \approx -0.29$, from their demand flexibility of ~3.5) with $\varepsilon_s \approx 0.10$, giving $M \approx 2.6$.

The gap between the two is driven almost entirely by the demand elasticity: staple food is far less substitutable than vegetable oil, which trades as a fungible complex. Reassuringly, the same equation reproduces both papers' headline results — Chen's 8.5% market share × 2.6 ≈ 22% (their estimate: 21.2%) and Roberts & Schlenker's 5% caloric shift × 6.7 ≈ 33%, within their reported 29–39% range. These checks are reproduced in code alongside the results.

Three modeling choices are worth flagging, and each is an explicit parameter in the computation below:

- **Market definition.** Because vegetable oils are close substitutes, the shock is best modeled in the aggregate oil complex (with palm as the marginal price), even though the demand originates in soy [@chen2026]. We therefore set $Q_0$ to total global vegetable-oil consumption, not just the volume currently used for biofuels.
- **Large-share breakdown.** The constant-elasticity approximation holds only for marginal shifts. As the ICCT demand grows past ~2032 it becomes a large fraction of the market, where the linear form overstates precision and the true price response is likely convex (accelerating) — consistent with the positively-skewed confidence interval in [@roberts2013] and the supply-insufficiency ICCT itself flags.
- **Short- vs. long-run.** Oil palm is perennial (~4-year lag to fruit), so short-run supply is nearly fixed (larger price spike) and the long-run response is larger [@chen2026]. The elasticities above are long-run values.
- **Soy co-product (meal) offset.** Crushing soybeans for oil jointly produces protein meal, whose revenue partially offsets the oil-price pressure — the vegetable-oil analogue of the distillers-grains adjustment in [@roberts2013], which scales their price effect by roughly two-thirds under full feed recycling. Chen et al. sidestep the issue by pricing the shock off palm, the marginal oil. We apply no co-product adjustment, so on this dimension our $\%\Delta P_t$ band should be read as an upper envelope.

The steps downstream of $\%\Delta P_t$ — converting the price shock into ILUC area and emissions — are handled by our own land-use regression, to be developed next. The derived price-shock series is exported to `data/derived/vegetable_oil_price_shock.csv` for that step.

## Data sources

### Projected demand for soy-based vegetable oil (from ICCT)

In the research brief
[Estimating the maximum potential vegetable oil demand for international shipping under the International Maritime Organization’s Net-Zero Framework](https://theicct.org/wp-content/uploads/2025/09/ID-450-Vegetable-Oil_research-brief_final.pdf) [@seber2025], ICCT estimates the projected demand for marine renewable biodiesel (and the associated demand for used cooking oil feedstock and virgin soy-based vegetable oil feedstock) under the proposed IMO Net-Zero Framework, from 2028 to 2035. This projected demand is summarized in the report's **Table 1** (page 5). We scraped the PDF report and turned this into a data tibble for use in our analysis:
```{r}
#| label: scrape-icct-table-1
#| message: false
#| warning: false

# Package versions are pinned via renv (see renv.lock); run renv::restore() to install.
# Functions are called with explicit pkg:: prefixes rather than attaching via library().

pdf_path <- file.path(
  "data",
  "icct_biofuel_demand_report",
  "ID-450-Vegetable-Oil_research-brief_final.pdf"
)

# Extract page 5 as text and keep only the data rows (years 2028–2035)
page5 <- pdftools::pdf_text(pdf_path)[5]
lines <- stringr::str_trim(stringr::str_split(page5, "\n")[[1]])
data_lines <- stringr::str_subset(lines, "^20(2[89]|3[0-5])\\b")

# Each data row yields exactly 10 numeric tokens:
#   year, shipping demand, base target, fossil LNG %,
#   UCO-RD %, UCO-RD bl, VO-RD %, VO-RD bl, UCO demand bl, VO demand bl
# Each year appears on its own line in the extracted text, so we
#pull the numeric tokens per row and split the two compound columns
# ("% energy demand / billion liters") into separate fields.
nums <- stringr::str_extract_all(data_lines, "[0-9]+\\.?[0-9]*")
mat <- do.call(rbind, lapply(nums, function(x) as.numeric(x[seq_len(10)])))

icct_table_1 <- tibble::as_tibble(mat, .name_repair = "minimal")
names(icct_table_1) <- c(
  "year",
  "shipping_energy_demand_ej",
  "base_target_trajectory_gco2e_mj",
  "fossil_lng_pct",
  "ucord_pct_energy_demand",
  "ucord_billion_liters",
  "vord_pct_energy_demand",
  "vord_billion_liters",
  "uco_demand_billion_liters",
  "vo_demand_billion_liters"
)

icct_table_1 <- icct_table_1 |>
  dplyr::mutate(year = as.integer(year)) |>
  dplyr::arrange(year)
```

#### Column schema

The scraped columns, their meaning, and units (derived from Table 1 and its
notes in the ICCT brief [@seber2025]). "UCO-RD" is used-cooking-oil-based
renewable diesel and "VO-RD" is vegetable-oil-based renewable diesel; the
*-RD* volumes are the fuel produced, while the *UCO/VO demand* columns are the
feedstock required to make it.

```{r}
#| label: icct-table-1-schema
#| message: false
#| warning: false

icct_table_1_schema <- tibble::tribble(
  ~column                           , ~description                                                                                     , ~units               ,
  "year"                            , "Calendar year of the projection"                                                                , "year (2028–2035)"   ,
  "shipping_energy_demand_ej"       , "Total international shipping energy demand (ships ≥5,000 GT; ICCT POLARIS model)"               , "EJ (exajoules)"     ,
  "base_target_trajectory_gco2e_mj" , "Base target GHG fuel-intensity trajectory the blended fuel must meet in the Net-Zero Framework" , "g CO2e/MJ"          ,
  "fossil_lng_pct"                  , "Fossil LNG share of shipping energy demand"                                                     , "% of energy demand" ,
  "ucord_pct_energy_demand"         , "UCO-RD share of shipping energy demand"                                                         , "% of energy demand" ,
  "ucord_billion_liters"            , "UCO-RD fuel volume blended into the fuel mix"                                                   , "billion liters"     ,
  "vord_pct_energy_demand"          , "VO-RD share of shipping energy demand"                                                          , "% of energy demand" ,
  "vord_billion_liters"             , "VO-RD fuel volume blended into the fuel mix"                                                    , "billion liters"     ,
  "uco_demand_billion_liters"       , "Used cooking oil (UCO) feedstock demand to produce the UCO-RD"                                  , "billion liters"     ,
  "vo_demand_billion_liters"        , "Virgin vegetable oil (VO) feedstock demand to produce the VO-RD"                                , "billion liters"
)

knitr::kable(
  icct_table_1_schema,
  col.names = c("Column", "Description", "Units")
)
```

```{r}
#| label: icct-table-1-interactive
#| message: false
#| warning: false

DT::datatable(
  icct_table_1,
  rownames = FALSE,
  filter = "top",
  caption = 'Caption from the ICCT report: "UCO-RD and VO-RD blend ratios and volumes to comply with the base target emissions trajectory. Notes: UCO-based renewable diesel availability is used as a constraint, and the rest of the demand is assumed to come from soy oil-based renewable diesel. UCO and VO demand to reach the estimated renewable diesel volumes are included. Estimations for shipping energy demand for ships of ≥ 5,000 gross tonnage are from the ICCT’s POLARIS Mode"',
  options = list(pageLength = 10, scrollX = TRUE)
)
```

The most important metric for our analysis is `vo_demand_billion_liters` (the projected soy-based virgin vegetable oil feedstock demand). It is this projected demand that we believe will most directly lead to a price shock that can be used in our land use model. Here we look at a time series of this projected demand:

```{r}
#| label: icct-demand-timeseries
#| message: false
#| warning: false

ts_data <- icct_table_1 |>
  dplyr::select(
    year,
    vo_demand_billion_liters
  )

dygraphs::dygraph(
  ts_data,
  main = "Projected demand for soy-based vegetable oil virgin feedstock",
  ylab = "Billion liters"
) |>
  dygraphs::dySeries("vo_demand_billion_liters", label = "VO demand") |>
  dygraphs::dyOptions(strokeWidth = 2, drawPoints = TRUE, pointSize = 3) |>
  dygraphs::dyLegend(show = "always") |>
  dygraphs::dyHighlight(highlightSeriesOpts = list(strokeWidth = 3))
```

## Results

### Derived price shock

Applying the demand-to-price mapping described in [Methods](#methods) to the ICCT projected demand series ([Data sources](#data-sources)) gives our first analytical output: the annual vegetable-oil price shock attributable to marine biofuel demand. Only two kinds of assumptions enter — the baseline market size $Q_0$ and the elasticity scenarios — and both are set explicitly here.

```{r}
#| label: price-shock-parameters
#| message: false
#| warning: false

# --- Parameters (see Methods for sourcing and caveats) ---

# Baseline global vegetable-oil consumption, billion liters.
# ~225 Mt (USDA, major vegetable oils, 2023/24) at ~0.92 kg/L.
q0_billion_liters <- 245

# Elasticity scenarios. eps_d is negative, so the multiplier denominator
# is eps_s + |eps_d|. The scenarios bracket the plausible demand response:
# vegetable oil trades as a fungible complex (more elastic demand, smaller
# price response) while staple food demand is very inelastic (larger one).
eps_s_chen <- 0.10 # long-run oil-crop supply elasticity (Chen et al. 2026)
eps_d_chen <- -0.29 # vegetable-oil demand elasticity (1 / 3.5 flexibility)
eps_s_rs <- 0.10 # staple-crop supply elasticity (Roberts & Schlenker 2013)
eps_d_rs <- -0.05 # staple-crop demand elasticity

m_chen <- 1 / (eps_s_chen - eps_d_chen) # ~2.6
m_rs <- 1 / (eps_s_rs - eps_d_rs) # ~6.7

scenario_params <- tibble::tribble(
  ~scenario, ~source,                                              ~eps_s,     ~eps_d,     ~multiplier,
  "Low",     "Chen, Sexton & Smith (2026): vegetable-oil complex", eps_s_chen, eps_d_chen, m_chen,
  "High",    "Roberts & Schlenker (2013): staple-food calories",   eps_s_rs,   eps_d_rs,   m_rs
)

knitr::kable(
  scenario_params,
  digits = 2,
  col.names = c(
    "Scenario", "Elasticity source",
    "Supply elasticity (ε_s)", "Demand elasticity (ε_d)", "Multiplier M"
  )
)
```

As a consistency check, each multiplier should reproduce its source paper's own headline price effect when fed that paper's demand shock:

```{r}
#| label: price-shock-validation
#| message: false
#| warning: false

validation <- tibble::tribble(
  ~check,                                                        ~shock_pct, ~m,     ~reported,
  "Chen et al.: BBD was 8.5% of vegetable-oil use, 2002-2018",   8.5,        m_chen, "21.2% (90% CI: 12.2-28.5%)",
  "Roberts & Schlenker: ethanol mandate = 5% of world calories", 5.0,        m_rs,   "~30% (range 29-39%)"
) |>
  dplyr::mutate(implied_pct = round(shock_pct * m, 1)) |>
  dplyr::select(check, shock_pct, implied_pct, reported)

knitr::kable(
  validation,
  col.names = c(
    "Validation case", "Demand shock (% of market)",
    "Implied %ΔP (our formula)", "Paper's reported %ΔP"
  )
)
```

Both implied values fall within the source papers' reported ranges. We now apply the multipliers to the ICCT demand series:

```{r}
#| label: derive-price-shock
#| message: false
#| warning: false

# %ΔP_t = (D_t / Q0) * M, expressed as a percent. The demand share is the
# outward demand shift as a fraction of the whole vegetable-oil market.
# beyond_marginal flags years where the shift exceeds ~25% of the market,
# outside the regime where a constant-elasticity approximation is reliable.
price_shock <- icct_table_1 |>
  dplyr::transmute(
    year,
    vo_demand_billion_liters,
    demand_share_pct = 100 * vo_demand_billion_liters / q0_billion_liters,
    pct_delta_p_low = demand_share_pct * m_chen,
    pct_delta_p_high = demand_share_pct * m_rs,
    beyond_marginal = demand_share_pct > 25
  )
```

```{r}
#| label: price-shock-table
#| message: false
#| warning: false

DT::datatable(
  price_shock |>
    dplyr::mutate(
      dplyr::across(where(is.numeric), \(x) round(x, 1)),
      beyond_marginal = ifelse(beyond_marginal, "yes", "")
    ),
  rownames = FALSE,
  colnames = c(
    "Year",
    "VO demand (billion L)",
    "Demand share of market (%)",
    "%ΔP low (Chen, M≈2.6)",
    "%ΔP high (Roberts-Schlenker, M≈6.7)",
    "Beyond marginal regime"
  ),
  caption = "Projected vegetable-oil price shock attributable to marine biofuel demand. The low/high columns bracket the price response using the two elasticity scenarios. Years flagged 'beyond marginal regime' (demand share > 25% of the market, from ~2032) exceed the range where the constant-elasticity approximation is reliable; treat those values as indicative of direction and scale, not precise magnitudes.",
  options = list(pageLength = 10, scrollX = TRUE, dom = "t")
)
```

```{r}
#| label: price-shock-timeseries
#| message: false
#| warning: false

dygraphs::dygraph(
  price_shock |> dplyr::select(year, pct_delta_p_low, pct_delta_p_high),
  main = "Projected vegetable-oil price shock from marine biofuel demand",
  ylab = "% change in vegetable oil price"
) |>
  dygraphs::dySeries(
    "pct_delta_p_high",
    label = "High (Roberts-Schlenker, M≈6.7)"
  ) |>
  dygraphs::dySeries("pct_delta_p_low", label = "Low (Chen, M≈2.6)") |>
  dygraphs::dyShading(from = 2031.5, to = 2035.5) |>
  dygraphs::dyOptions(strokeWidth = 2, drawPoints = TRUE, pointSize = 3) |>
  dygraphs::dyLegend(show = "always") |>
  dygraphs::dyHighlight(highlightSeriesOpts = list(strokeWidth = 3))
```

The shaded region (2032 onward) marks years where marine demand exceeds ~25% of the baseline vegetable-oil market — beyond the marginal regime, where the true price response is likely convex and these linear values understate it. Finally, we export the derived series for the downstream land-use step:

```{r}
#| label: export-price-shock
#| message: false
#| warning: false

# Hand-off to the price -> ILUC regression (full precision, no rounding).
dir.create(file.path("data", "derived"), showWarnings = FALSE, recursive = TRUE)
utils::write.csv(
  price_shock,
  file.path("data", "derived", "vegetable_oil_price_shock.csv"),
  row.names = FALSE
)
```

The exported series (`pct_delta_p_low`, `pct_delta_p_high` in `data/derived/vegetable_oil_price_shock.csv`) is the input to the land-use step — converting the price shock into ILUC area and emissions — which we will develop next.

## Discussion

*Discussion of findings will go here.*

## References

::: {#refs}
:::
