---
title: "Assessing the ILUC implications of increasing marine biofuel demand"
subtitle: "Prototype Proof-of-Concept Analysis"
authors: Environmental Markets Lab (emLab) at University of California, Santa Barbara (UCSB). For International Council on Clean Transportation (ICCT)
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
    embed-resources: false
bibliography: references.bib
---

## Overview

This report documents a prototype proof-of-concept analysis investigating the projected indirect land use change (ILUC) implications of the increasing marine biofuel demand that could be incentivized by the yet-to-be-approved IMO Net-Zero Framework. This prototype analysis is being developed by Environmental Markets Lab (emLab) at University of California, Santa Barbara (UCSB) for International Council on Clean Transportation (ICCT). All results presented in this analysis should be considered exploratory and preliminary.

## Background

### The IMO Net-Zero Framework

At its 83rd session (April 2025), the International Maritime Organization (IMO) approved draft amendments to MARPOL Annex VI that, if adopted, would establish a legally binding Net-Zero Framework to decarbonize international shipping, with the goal of net-zero GHG emissions by or around 2050 [@seber2025; @sturrup2026]. The framework applies to all ships of 5,000 gross tonnage and above on international voyages. Each year, a ship reports its *attained* greenhouse-gas fuel intensity (GFI) — the average well-to-wake life-cycle emissions per unit of energy used on board, in grams of CO₂e per megajoule (g CO₂e/MJ) — and compares it against annual targets. A formal adoption vote slated for October 2025 was delayed by at least a year, so compliance may begin in 2029 rather than 2028.

**Two-tier targets.** The framework sets two GFI targets, both defined as reductions from the 2008 fleet-average baseline of 93.3 g CO₂e/MJ, and both tightening every year:

- A **Base target** (the less stringent tier), falling from a 4% reduction in 2028 (89.6 g CO₂e/MJ) to a 30% reduction by 2035 (65.3 g CO₂e/MJ), and set to reach a 65% reduction by 2040.
- A stricter **Direct Compliance target**, falling from a 17% reduction in 2028 (77.4 g CO₂e/MJ) to a 43% reduction by 2035 (53.2 g CO₂e/MJ).

**Carbon pricing.** A ship's position relative to these two thresholds determines what it pays for each tonne of excess CO₂e (a "compliance balance"):

- Above the Base target → **Tier 2** penalty of **\$380/t CO₂e** (remedial units); banked surplus units from over-complying ships may be used instead.
- Between the Direct Compliance and Base targets → reduced **Tier 1** penalty of **\$100/t CO₂e**, with no option to offset with surplus units.
- Below the Direct Compliance target → the ship over-complies and earns tradable surplus units.

**Fuel carbon intensities.** Because the framework judges fuels only on their life-cycle GFI, the compliance value of each fuel is set by how far its carbon intensity sits below the fossil incumbents. The values ICCT uses [@seber2025] illustrate the gap that makes crop-based biofuels attractive:

| Fuel | Life-cycle carbon intensity (g CO₂e/MJ) |
|---|---|
| 2008 baseline | 93.3 |
| Heavy fuel oil (HFO) | 92.8 |
| Fossil LNG | 85.7 |
| Soy vegetable-oil renewable diesel (VO-RD) | 42.3 |
| Soy FAME biodiesel | 35.7 |
| Used-cooking-oil renewable diesel (UCO-RD) | 13.1 |

: Life-cycle (well-to-wake) carbon intensities used in the ICCT demand brief [@seber2025].

**Why the framework incentivizes biofuel demand.** The framework contains no eligibility criteria for compliance fuels and evaluates them solely on life-cycle emissions; crucially, the underlying LCA guidelines are still in development and include no safeguards for land-use change or other indirect emissions. Blending low-carbon-intensity biofuel into an otherwise fossil fuel mix lowers a ship's attained GFI, and as long as the cost of that blending stays below the \$380/t Tier 2 penalty, the cost-minimizing strategy is to blend just enough biofuel to meet the Base target and pay only the cheaper \$100/t Tier 1 charge. ICCT's cost-optimization finds the cheapest pathway — used-cooking-oil renewable diesel — is quickly exhausted by limited supply (~13.6 billion L of fuel, about 5% of shipping energy demand), so virgin vegetable-oil renewable diesel absorbs the remaining compliance burden, growing to 46% of shipping energy demand by 2035 [@seber2025]. This is the demand shock our analysis traces.

### The indirect land-use change concern

Indirect land use change (ILUC) occurs when demand for crop-based biofuel feedstocks raises agricultural commodity prices, and higher prices push cropland expansion into forest and other carbon-rich land elsewhere in the world. These emissions fall outside the fuel's direct supply chain, so lifecycle accounting that stops at the refinery gate misses them, even though they can rival or exceed the emissions of the fossil fuel being displaced. Because the current draft of the Net-Zero Framework contains no ILUC safeguards, a fuel mix that is fully compliant on paper could carry substantial uncounted emissions — making this question directly relevant to marine fuel policy.

Our analysis traces this chain for marine biofuels:

```{mermaid}
flowchart LR
  A["IMO Net-Zero Framework<br/><i>policy</i>"] --> F["Marine renewable-diesel<br/>(VO-RD) fuel demand<br/><i>ICCT cost-optimization</i>"]
  F --> B["Virgin vegetable-oil<br/>feedstock demand D<br/><i>1.22 kg oil per kg RD</i>"]
  B --> C["Vegetable-oil<br/>price shock %ΔP<br/><i>this report</i>"]
  C --> D["ILUC area and<br/>carbon emissions<br/><i>our land-use model, next</i>"]
```

### Literature review

We draw on four primary sources, each covering a different link in the chain. Two further sources on how price shocks transmit across the different vegetable oils [@cui2017; @santeramo2017] are discussed in [Methods](#a-common-price-shock-across-oils).

| Source | Role in our pipeline | Key values we use |
|---|---|---|
| ICCT demand brief [@seber2025] | Projects the demand shock $D_t$ | 0 → 139 billion L virgin VO, 2028–2035 |
| ICCT climate & revenue brief [@sturrup2026] | Benchmark fixed-factor translation of $D_t$ into ILUC emissions | 62 g CO₂e/MJ factor; ~950 Mt CO₂e cumulative |
| Chen, Sexton & Smith [@chen2026] | Empirical demand → price → deforestation chain (the design we emulate) | Demand flexibility $-3.5$; price multiplier $M \approx 2.5$; ILUC CI 84.1 g CO₂e/MJ |
| Roberts & Schlenker [@roberts2013] | Identification strategy and elasticities for the demand → price step | $\varepsilon_s \approx 0.10$, $\varepsilon_d \approx -0.05$; $M \approx 6.7$ |

: Roles of the four sources in our analysis pipeline.

#### Estimating the maximum potential vegetable oil demand for international shipping under the IMO Net-Zero Framework [@seber2025]

This brief estimates how much vegetable oil demand the Net-Zero Framework could induce. It does not quantify the resulting ILUC emissions, but it flags the risk and provides the demand projection that our analysis and the companion brief below both build on.

- **Policy mechanism.** The framework sets two annual greenhouse-gas fuel-intensity (GFI) targets against a 2008 baseline of 93.3 g CO₂e/MJ: a Base target, with a \$380/t CO₂e Tier 2 penalty for exceeding it, and a stricter Direct Compliance target, with a \$100/t Tier 1 penalty between the two. Fuels are judged only on their direct life-cycle emissions; there are no ILUC safeguards, so crop-based biofuels are an attractive compliance option.
- **Demand derivation.** ICCT builds a least-cost sector-wide fuel mix for 2028–2035. Fossil LNG enters exogenously (0.81 EJ, about 8% of energy demand by 2035, from their POLARIS model [@polaris2024]), and biofuels are then blended until the fleet-wide average carbon intensity just meets the Base target. Since blending costs less than \$380/t, the cost-minimizing strategy is to meet the Base target and pay only the Tier 1 penalty.
- **Feedstock logic.** The cheapest pathway, used-cooking-oil renewable diesel (UCO-RD, 13.1 g CO₂e/MJ), is blended first up to its supply limit; virgin vegetable-oil renewable diesel (soy-RD, 42.3 g CO₂e/MJ) fills the rest. UCO supply is the binding constraint — roughly 14.3 billion L of feedstock, or 13.6 billion L of fuel, covering only about 5% of shipping energy demand — so soy oil absorbs the remaining compliance burden.
- **Key finding.** VO-RD grows to 46% of shipping energy demand by 2035, implying virgin vegetable-oil demand of up to 139 billion liters (Table 1, p. 5), roughly 3.5 times the 40 billion liters of vegetable oil used for biofuels today. ICCT judges existing virgin oil supply insufficient to meet the targets beyond 2032 without new cropland or diversion from other sectors.
- **Use in this analysis.** We scrape Table 1 and use the `vo_demand_billion_liters` column as the demand shock for our land-use model (see [Data sources](#data-sources)).

#### Climate and revenue considerations for the IMO Net-Zero Framework [@sturrup2026]

This companion brief takes the demand projection from @seber2025 and translates it into ILUC emissions — the same step our prototype performs, using different methods. It covers two topics: biofuel use for compliance, and the climate finance raised from penalties.

- **Approach.** They assume the projected vegetable-oil demand is met with soy-RD and apply a single illustrative ILUC emission factor of 62 g CO₂e/MJ from the U.S. EPA's Model Comparison Exercise [@epa2023mce]. Since the blend is assumed to meet the Base target, the ILUC-adjusted carbon intensity of the fuel mix is simply the Base-target GFI plus the ILUC contribution.
- **Provenance of the 62.** The value is the GCAM estimate of land-use-change carbon intensity for soybean oil biodiesel in Table 7.7-1 of the EPA exercise; the other models report 295 (ADAGE), 10 (GTAP), 23 (GLOBIOM), and 10 (GREET). The exercise ran a 1 billion gallon per year soybean-biodiesel shock through five models, and the spread is a fair measure of how unsettled ILUC estimates remain. One unit caveat: the EPA table is in kg CO₂e/MMBTU (62 kg CO₂e/MMBTU ≈ 59 g CO₂e/MJ), while the ICCT brief cites the figure as 62 g CO₂e/MJ.
- **Caveat on the factor.** The value may be high for U.S. soy but likely understates ILUC for Indonesian palm or Brazilian soy; GLOBIOM-based estimates for the EU exceed 100 g CO₂e/MJ for high-risk feedstocks [@valin2015]. It is an illustrative point estimate, not a bound.
- **ILUC findings.** The ILUC contribution to fleet GFI grows from 0.9 g CO₂e/MJ in 2029 to 28.5 by 2035 (their Table A1), pushing the ILUC-adjusted GFI to 93.9 g CO₂e/MJ — above the 93.3 fossil baseline. A compliant fuel mix could be worse for the climate than the fuel it replaces.
- **Magnitude.** Annual ILUC emissions reach roughly 285 Mt CO₂e by 2035; the 2028–2035 cumulative total of about 950 Mt CO₂e is comparable to a full year of well-to-wake emissions from international shipping (939 Mt CO₂e in 2024).
- **Revenue interaction.** Blending biofuels to avoid Tier 2 penalties erodes Net-Zero Fund revenue (projected at \$11–12 billion per year), and raising the Tier 1 price would make blending more attractive still, cutting revenue further while worsening ILUC.
- **Policy conclusion.** They recommend ILUC accounting in the IMO LCA guidelines (as in CORSIA and California's LCFS), limits on food- and feed-based biofuels, and financial incentives reserved for advanced fuels.

#### Using vegetable oils for biofuel accelerates tropical deforestation and increases carbon emissions [@chen2026]

This working paper is the closest empirical analog to our prototype. Rather than assuming an ILUC factor, the authors estimate the full demand → price → land conversion → emissions chain econometrically; our analysis aims to reproduce that structure for the marine case.

- **Method.** Three steps: demand and supply models for the major oil crops (palm, soybean, rapeseed, sunflower) establish how biomass-based diesel (BBD) demand moves palm prices; a panel regression on 32,394 grid cells (9 km × 9 km, 2002–2018) of remote-sensing data estimates the effect of palm prices on deforestation, with fixed effects plus reverse-causality and weather checks; and spatial biomass, peatland, and forest-type data convert deforested area into carbon emissions.
- **Price response.** BBD averaged 8.5% of global vegetable-oil consumption over 2002–2018, which the authors estimate raised palm oil prices 21.2% over business-as-usual (90% CI: 12.2–28.5%). The key parameter is a demand flexibility of 3.5: a 1% drop in oil-crop production raises palm prices 3.5%.
- **Supply response.** Short-run palm supply elasticity is small and insignificant (palm is a perennial that takes about four years to fruit); the long-run elasticity is about 0.1. Some 84% of the competing-crop supply response comes from area expansion and only 16% from yield, at odds with the strong intensification response many CGE-based ILUC models assume.
- **Deforestation.** A \$1,000/ha increase in attainable palm revenue drives about 9.3 additional hectares of deforestation per average grid cell, which aggregates to 1.7 million ha of forest-to-palm conversion (90% CI: 1.0–2.3 Mha) — about 20% of all such conversion in Indonesia and Malaysia over the period. Conversion falls mostly on natural forest and is driven mainly by industrial growers.
- **Emissions.** Net land-use-change emissions are 1,086 Tg CO₂e (biomass loss 758, peat 614, soil carbon 95, less 380 sequestered by the palms), amortized over 25 years. That works out to an ILUC carbon intensity of 84.1 g CO₂e/MJ (90% CI: 48.7–113.5); with cultivation, processing, and transport added (34.6–63.5), total carbon intensity is 118.7–147.6 g CO₂e/MJ, or 22–52% above petroleum diesel (~97). The authors treat this as a conservative lower bound, since it excludes ILUC from other oils and from palm outside the study region.
- **Substitution.** Vegetable oils are close substitutes, so policies that restrict only palm still bid up palm prices by diverting demand to soy, rapeseed, and sunflower. This is the same market linkage behind the soy-driven ILUC concern in the ICCT briefs, and it is why the authors favor waste feedstocks (UCO, tallow) that avoid land-use effects.

#### Identifying supply and demand elasticities of agricultural commodities: implications for the US ethanol mandate [@roberts2013]

This paper supplies the identification strategy and elasticity estimates for the demand-to-price step: how a mandated biofuel demand shock translates into an equilibrium commodity price change.

- **Identification.** The paper uses competitive storage theory to justify weather-driven yield shocks as instruments: past shocks shift inventories and futures prices, identifying supply, while contemporaneous shocks move current prices, identifying demand. First-stage F-statistics exceed 10. The analysis covers corn, wheat, rice, and soybeans (about 75% of global food calories), aggregated into caloric equivalents.
- **Elasticities.** Supply is 0.09–0.12 and demand −0.03 to −0.07 (their Table 1), both larger in magnitude than uninstrumented estimates. The implied price multiplier, 1/(β~s~ − β~d~), is 5.8–7.7: small demand shifts produce large price changes.
- **Extensive margin.** The land-area elasticity is 0.07–0.08, with no significant fertilizer response. Added production comes from expanding cropland rather than intensification, which is the mechanism connecting biofuel demand to land-use change.
- **Ethanol-mandate application.** Diverting about 5% of world caloric production (roughly one-third of US corn) to ethanol raises staple food prices about 30% (95% CI: 14–35%; closer to 20% if a third of the calories return as feed byproduct), expands world crop area about 2%, and implies a consumer-surplus loss near \$180 billion per year, borne mostly by net-food-importing developing countries.
- **Use in this analysis.** We take the elasticity estimates and the multiplier logic from this paper to map the marine demand shock into a vegetable-oil price change; land conversion then follows through the mechanism in @chen2026.

## Methods

### From demand projection to a vegetable-oil price shock

Our land-use model takes as input a price shock: the proportional change in the vegetable-oil price attributable to new marine biofuel demand. We derive it from the ICCT demand series (`vo_demand_billion_liters`; see [Data sources](#data-sources)) using the standard comparative-statics result that an outward demand shift moves the equilibrium price by the size of the shift divided by the gap between the supply and demand elasticities [@roberts2013]:

$$ \%\Delta P_t \;\approx\; \underbrace{\frac{D_t}{Q_0}}_{\text{demand shift}} \;\times\; \underbrace{\frac{1}{\varepsilon_s - \varepsilon_d}}_{\text{price multiplier } M} $$

where $D_t$ is the projected marine virgin vegetable-oil demand in year $t$ (billion L), $Q_0$ is baseline global vegetable-oil consumption (billion L), $\varepsilon_s > 0$ is the supply elasticity, and $\varepsilon_d < 0$ is the demand elasticity (so the denominator is $\varepsilon_s + |\varepsilon_d|$). The shock depends on demand as a share of the whole market, not on absolute volume.

Our **central estimate** takes both elasticities from Chen, Sexton & Smith [@chen2026], since theirs is the empirical setting closest to ours (the vegetable-oil complex rather than staple-food calories):

- **Demand.** The baseline demand flexibility from their Figure 2, panel A (row 1; Table S1, column 3) is $-3.523$ (90% CI: $[-5.3, -1.7]$): a 1% drop in aggregate oil-crop production raises palm oil prices 3.5%. Inverting gives $\varepsilon_d = 1/{-3.523} \approx -0.284$.
- **Supply.** Their long-run supply elasticities with respect to the palm price are ~0.10 for palm and ~0.13 for the competing oil crops, which they combine as a production-share-weighted average, $\varepsilon_s \approx 0.12$.

Together these give a central multiplier $M = 1/(\varepsilon_s - \varepsilon_d) \approx 2.5$ — exactly the machinery Chen et al. use for their own counterfactual ($\%\Delta P = \%\Delta Q / (\varepsilon_s - \varepsilon_d)$). As a **sensitivity**, we retain the Roberts & Schlenker [@roberts2013] staple-food elasticities ($\varepsilon_s \approx 0.10$, $\varepsilon_d \approx -0.05$), whose much less elastic demand gives $M \approx 6.7$. As a check on the formula and our parameter choices, it recovers both papers' headline results: 8.5% × 2.5 ≈ 21% against Chen's reported 21.2%, and 5% × 6.7 ≈ 33% against Roberts & Schlenker's reported 29–39%. We repeat these checks in code with the results.

A few modeling choices matter for interpretation:

- **Market definition.** Because vegetable oils are close substitutes, the shock plays out in the aggregate oil complex, with palm as the marginal price, even though the marine demand lands on soy [@chen2026]. We therefore set $Q_0$ to match the market over which the Chen elasticities are defined: global production of the four major oil crops (palm, soybean, rapeseed, sunflower). USDA's world supply-and-distribution tables put 2024/25 production of the four oils at 200.8 Mt (palm 78.3, soybean 68.6, rapeseed 33.9, sunflowerseed 20.1) [@usda2025oilseeds], about 218 billion liters at ~0.92 kg/L. (USDA's total across all nine tracked vegetable oils is ~231 Mt; using the four-oil complex keeps $Q_0$ consistent with the demand flexibility's denominator.)
- **Single demand year.** The land-use model takes one price shock, not a time series. We use the final year of the ICCT projection (2035, virgin VO demand of 139 billion L) — the mature-policy year, when the framework's targets have tightened enough for VO-RD to carry most of the compliance burden. We still report the full 2028–2035 series for context.
- **Large shocks.** The constant-elasticity approximation is a marginal result. From about 2032 the ICCT demand becomes a large fraction of the market, where the linear form loses reliability and the true price response is likely convex, consistent with the positively skewed confidence interval in @roberts2013 and with the supply shortfall ICCT itself flags.
- **Short- vs. long-run.** Oil palm is a perennial with roughly a four-year lag to fruit, so short-run supply is nearly fixed and the long-run response is larger [@chen2026]. The elasticities above are long-run values.
- **Soy co-product.** Crushing soybeans for oil also produces protein meal, and meal revenue partially offsets the oil-price pressure — the analogue of the distillers-grains adjustment in @roberts2013, which scales their price effect by roughly two-thirds under full feed recycling. Chen et al. avoid the issue by pricing the shock off palm. We apply no adjustment, so on this margin our estimates err high.

### A common price shock across oils {#a-common-price-shock-across-oils}

The formula above yields a single percentage price shock, but our land-use model takes crop-specific price matrices. Following Chen, Sexton & Smith [@chen2026], we apply the **same percentage shock to each vegetable-oil crop price** (palm, soy, canola). Their model carries only one price — everything is denominated in the palm price, and a demand shock landing on any oil is expressed as a share of the combined four-oil quantity — and they support this "one market, one price" treatment empirically: palm prices respond to production shocks in competing oil crops about as strongly as to palm's own (their Figure 2A), and the four oil prices are cointegrated with near-unit long-run coefficients (soybean 1.04, canola 1.01, sunflower 0.86; their Table S5), meaning proportional shocks transmit roughly one-for-one across oils.

Studies that model the oils as separate markets find transmission that is real but partial and heterogeneous — Cui & Martin [@cui2017] and an ICCT consultant report on cross-price elasticities [@santeramo2017] — so relaxing the uniform-shock assumption with oil-specific pass-through is a potential refinement for future work.

### From price shock to ILUC carbon intensity

Converting $\%\Delta P$ into ILUC area and emissions uses our land-use model (developed separately), which regresses land conversion on crop prices and simulates forward by country. The hand-off works as follows:

1. **Two scenarios.** A business-as-usual run with baseline crop prices, and a marine-demand run in which the vegetable-oil crop prices (palm, soy, canola) are raised by the common percentage shock derived here. The model's carbon price is set to zero in both scenarios, so the difference between runs isolates the price-driven land-use response.
2. **Differencing and annualization.** Differencing the two runs gives the additional land-use-change emissions attributable to marine biofuel demand, annualized over the 150-year simulation horizon.
3. **Carbon intensity.** Dividing annualized emissions by the annual fuel energy delivered (MJ of VO-RD in the 2035 demand year) yields an ILUC carbon intensity in g CO₂e/MJ, directly comparable to — and addable to — the direct carbon intensity ICCT assumes for soy VO-RD (42.3 g CO₂e/MJ [@seber2025]).
4. **Country decomposition.** Because the model resolves land conversion by country, the ILUC intensity decomposes into country contributions (e.g., a stacked bar chart), showing where the induced conversion falls.

The 2035 price shock is written to `data/derived/vegetable_oil_price_shock_2035.csv` (and the full series to `data/derived/vegetable_oil_price_shock.csv`) for that step.

## Data sources

### Projected demand for soy-based vegetable oil (from ICCT)

In the research brief
[Estimating the maximum potential vegetable oil demand for international shipping under the International Maritime Organization’s Net-Zero Framework](https://theicct.org/wp-content/uploads/2025/09/ID-450-Vegetable-Oil_research-brief_final.pdf) [@seber2025], ICCT estimates the projected demand for marine renewable biodiesel (and the associated demand for used cooking oil feedstock and virgin soy-based vegetable oil feedstock) under the proposed IMO Net-Zero Framework, from 2028 to 2035. This projected demand is summarized in the report's **Table 1** (page 5). We scraped the PDF report and turned this into a data tibble for use in our analysis:
```{r}
#| label: scrape-icct-table-1
#| message: false
#| warning: false

# Package versions are pinned via renv (see renv.lock); run renv::restore() to install.
# Functions are called with explicit pkg:: prefixes rather than attaching via library().

pdf_path <- file.path(
  "data",
  "icct_biofuel_demand_report",
  "ID-450-Vegetable-Oil_research-brief_final.pdf"
)

# Extract page 5 as text and keep only the data rows (years 2028–2035)
page5 <- pdftools::pdf_text(pdf_path)[5]
lines <- stringr::str_trim(stringr::str_split(page5, "\n")[[1]])
data_lines <- stringr::str_subset(lines, "^20(2[89]|3[0-5])\\b")

# Each data row yields exactly 10 numeric tokens:
#   year, shipping demand, base target, fossil LNG %,
#   UCO-RD %, UCO-RD bl, VO-RD %, VO-RD bl, UCO demand bl, VO demand bl
# Each year appears on its own line in the extracted text, so we
#pull the numeric tokens per row and split the two compound columns
# ("% energy demand / billion liters") into separate fields.
nums <- stringr::str_extract_all(data_lines, "[0-9]+\\.?[0-9]*")
mat <- do.call(rbind, lapply(nums, function(x) as.numeric(x[seq_len(10)])))

icct_table_1 <- tibble::as_tibble(mat, .name_repair = "minimal")
names(icct_table_1) <- c(
  "year",
  "shipping_energy_demand_ej",
  "base_target_trajectory_gco2e_mj",
  "fossil_lng_pct",
  "ucord_pct_energy_demand",
  "ucord_billion_liters",
  "vord_pct_energy_demand",
  "vord_billion_liters",
  "uco_demand_billion_liters",
  "vo_demand_billion_liters"
)

icct_table_1 <- icct_table_1 |>
  dplyr::mutate(year = as.integer(year)) |>
  dplyr::arrange(year)
```

#### Column schema

The scraped columns, their meaning, and units (derived from Table 1 and its
notes in the ICCT brief [@seber2025]). "UCO-RD" is used-cooking-oil-based
renewable diesel and "VO-RD" is vegetable-oil-based renewable diesel; the
*-RD* volumes are the fuel produced, while the *UCO/VO demand* columns are the
feedstock required to make it.

```{r}
#| label: icct-table-1-schema
#| message: false
#| warning: false

icct_table_1_schema <- tibble::tribble(
  ~column                           , ~description                                                                                     , ~units               ,
  "year"                            , "Calendar year of the projection"                                                                , "year (2028–2035)"   ,
  "shipping_energy_demand_ej"       , "Total international shipping energy demand (ships ≥5,000 GT; ICCT POLARIS model)"               , "EJ (exajoules)"     ,
  "base_target_trajectory_gco2e_mj" , "Base target GHG fuel-intensity trajectory the blended fuel must meet in the Net-Zero Framework" , "g CO2e/MJ"          ,
  "fossil_lng_pct"                  , "Fossil LNG share of shipping energy demand"                                                     , "% of energy demand" ,
  "ucord_pct_energy_demand"         , "UCO-RD share of shipping energy demand"                                                         , "% of energy demand" ,
  "ucord_billion_liters"            , "UCO-RD fuel volume blended into the fuel mix"                                                   , "billion liters"     ,
  "vord_pct_energy_demand"          , "VO-RD share of shipping energy demand"                                                          , "% of energy demand" ,
  "vord_billion_liters"             , "VO-RD fuel volume blended into the fuel mix"                                                    , "billion liters"     ,
  "uco_demand_billion_liters"       , "Used cooking oil (UCO) feedstock demand to produce the UCO-RD"                                  , "billion liters"     ,
  "vo_demand_billion_liters"        , "Virgin vegetable oil (VO) feedstock demand to produce the VO-RD"                                , "billion liters"
)

knitr::kable(
  icct_table_1_schema,
  col.names = c("Column", "Description", "Units")
)
```

```{r}
#| label: icct-table-1-interactive
#| message: false
#| warning: false

DT::datatable(
  icct_table_1,
  rownames = FALSE,
  filter = "top",
  caption = 'Caption from the ICCT report: "UCO-RD and VO-RD blend ratios and volumes to comply with the base target emissions trajectory. Notes: UCO-based renewable diesel availability is used as a constraint, and the rest of the demand is assumed to come from soy oil-based renewable diesel. UCO and VO demand to reach the estimated renewable diesel volumes are included. Estimations for shipping energy demand for ships of ≥ 5,000 gross tonnage are from the ICCT’s POLARIS Mode"',
  options = list(pageLength = 10, scrollX = TRUE)
)
```

The most important metric for our analysis is `vo_demand_billion_liters` (the projected soy-based virgin vegetable oil feedstock demand). It is this projected demand that we believe will most directly lead to a price shock that can be used in our land use model. Here we look at a time series of this projected demand:

```{r}
#| label: icct-demand-timeseries
#| message: false
#| warning: false

ts_data <- icct_table_1 |>
  dplyr::select(
    year,
    vo_demand_billion_liters
  )

dygraphs::dygraph(
  ts_data,
  main = "Projected demand for soy-based vegetable oil virgin feedstock",
  ylab = "Billion liters"
) |>
  dygraphs::dySeries("vo_demand_billion_liters", label = "VO demand") |>
  dygraphs::dyOptions(strokeWidth = 2, drawPoints = TRUE, pointSize = 3) |>
  dygraphs::dyLegend(show = "always") |>
  dygraphs::dyHighlight(highlightSeriesOpts = list(strokeWidth = 3))
```

## Results

### Derived price shock

We now apply the mapping from [Methods](#methods) to the ICCT demand series to estimate the annual vegetable-oil price shock attributable to marine biofuel demand. The only assumptions are the baseline market size $Q_0$ and the elasticity scenarios, set explicitly below.

```{r}
#| label: price-shock-parameters
#| message: false
#| warning: false

# --- Parameters (see Methods for sourcing and caveats) ---

# Baseline market size Q0: 2024/25 world production of the four major oil
# crops over which the Chen et al. (2026) elasticities are defined
# (USDA FAS, Oilseeds: World Markets and Trade, May 2025, world S&D
# tables 16-19), converted to billion liters at ~0.92 kg/L.
production_mt <- c(palm = 78.25, soybean = 68.57, rapeseed = 33.90, sunflower = 20.08)
q0_mt <- sum(production_mt) # 200.8 Mt
oil_density_kg_l <- 0.92
q0_billion_liters <- q0_mt / oil_density_kg_l # ~218 billion L

# Central elasticities (Chen et al. 2026). Demand: baseline demand
# flexibility of -3.523 (Figure 2A row 1 / Table S1 col 3; 90% CI
# [-5.3, -1.7]) inverts to eps_d = -0.284. Supply: long-run elasticities
# w.r.t. the palm price of ~0.10 (palm) and ~0.13 (competing oil crops),
# combined as a production-share-weighted average per their eps_S formula.
demand_flexibility_chen <- -3.523
eps_d_chen <- 1 / demand_flexibility_chen
share_palm <- production_mt[["palm"]] / q0_mt
eps_s_chen <- 0.10 * share_palm + 0.13 * (1 - share_palm)

# Sensitivity: Roberts & Schlenker (2013) staple-food elasticities, whose
# much less elastic demand gives a larger multiplier.
eps_s_rs <- 0.10
eps_d_rs <- -0.05

# eps_d is negative, so the multiplier denominator is eps_s + |eps_d|.
m_chen <- 1 / (eps_s_chen - eps_d_chen) # ~2.5
m_rs <- 1 / (eps_s_rs - eps_d_rs) # ~6.7

scenario_params <- tibble::tribble(
  ~scenario     , ~source                                              , ~eps_s     , ~eps_d     , ~multiplier ,
  "Central"     , "Chen, Sexton & Smith (2026): vegetable-oil complex" , eps_s_chen , eps_d_chen , m_chen      ,
  "Sensitivity" , "Roberts & Schlenker (2013): staple-food calories"   , eps_s_rs   , eps_d_rs   , m_rs
)

knitr::kable(
  scenario_params,
  digits = 2,
  col.names = c(
    "Scenario",
    "Elasticity source",
    "Supply elasticity (ε_s)",
    "Demand elasticity (ε_d)",
    "Multiplier M"
  )
)
```

Before applying the multipliers to our series, we check that each one recovers its source paper's own headline price effect:

```{r}
#| label: price-shock-validation
#| message: false
#| warning: false

validation <- tibble::tribble(
  ~check                                                        , ~shock_pct , ~m     , ~reported                    ,
  "Chen et al.: BBD was 8.5% of vegetable-oil use, 2002-2018"   , 8.5        , m_chen , "21.2% (90% CI: 12.2-28.5%)" ,
  "Roberts & Schlenker: ethanol mandate = 5% of world calories" , 5.0        , m_rs   , "~30% (range 29-39%)"
) |>
  dplyr::mutate(implied_pct = round(shock_pct * m, 1)) |>
  dplyr::select(check, shock_pct, implied_pct, reported)

knitr::kable(
  validation,
  col.names = c(
    "Validation case",
    "Demand shock (% of market)",
    "Implied %ΔP (our formula)",
    "Paper's reported %ΔP"
  )
)
```

Both checks land within the papers' reported ranges. Applying the multipliers to the ICCT series:

```{r}
#| label: derive-price-shock
#| message: false
#| warning: false

# %ΔP_t = (D_t / Q0) * M, expressed as a percent. The demand share is the
# outward demand shift as a fraction of the whole vegetable-oil market.
# beyond_marginal flags years where the shift exceeds ~25% of the market,
# outside the regime where a constant-elasticity approximation is reliable.
price_shock <- icct_table_1 |>
  dplyr::transmute(
    year,
    vo_demand_billion_liters,
    demand_share_pct = 100 * vo_demand_billion_liters / q0_billion_liters,
    pct_delta_p_central = demand_share_pct * m_chen,
    pct_delta_p_high = demand_share_pct * m_rs,
    beyond_marginal = demand_share_pct > 25
  )
```

```{r}
#| label: price-shock-table
#| message: false
#| warning: false

DT::datatable(
  price_shock |>
    dplyr::mutate(
      dplyr::across(where(is.numeric), \(x) round(x, 1)),
      beyond_marginal = ifelse(beyond_marginal, "yes", "")
    ),
  rownames = FALSE,
  colnames = c(
    "Year",
    "VO demand (billion L)",
    "Demand share of market (%)",
    "%ΔP central (Chen, M≈2.5)",
    "%ΔP sensitivity (Roberts-Schlenker, M≈6.7)",
    "Beyond marginal regime"
  ),
  caption = "Projected vegetable-oil price shock attributable to marine biofuel demand. The central column uses the Chen et al. (2026) elasticities; the sensitivity column uses Roberts & Schlenker (2013). Years flagged 'beyond marginal regime' (demand share > 25% of the market, from ~2032) exceed the range where the constant-elasticity approximation is reliable; treat those values as indicative of direction and scale, not precise magnitudes.",
  options = list(pageLength = 10, scrollX = TRUE, dom = "t")
)
```

```{r}
#| label: price-shock-timeseries
#| message: false
#| warning: false

dygraphs::dygraph(
  price_shock |> dplyr::select(year, pct_delta_p_central, pct_delta_p_high),
  main = "Projected vegetable-oil price shock from marine biofuel demand",
  ylab = "% change in vegetable oil price"
) |>
  dygraphs::dySeries(
    "pct_delta_p_high",
    label = "Sensitivity (Roberts-Schlenker, M≈6.7)"
  ) |>
  dygraphs::dySeries("pct_delta_p_central", label = "Central (Chen, M≈2.5)") |>
  dygraphs::dyOptions(strokeWidth = 2, drawPoints = TRUE, pointSize = 3) |>
  dygraphs::dyLegend(show = "always") |>
  dygraphs::dyHighlight(highlightSeriesOpts = list(strokeWidth = 3))
```

From about 2032 on, marine demand becomes a large fraction of the baseline market, where the constant-elasticity approximation is unreliable and, if anything, understates the true response.

### Hand-off to the land-use model

The land-use model takes a single price shock: the central (Chen) estimate for 2035, the final year of the ICCT projection (see [Methods](#a-common-price-shock-across-oils)). This same percentage is applied to each vegetable-oil crop price (palm, soy, canola) in the marine-demand scenario's price matrix.

```{r}
#| label: export-price-shock
#| message: false
#| warning: false

# Hand-off to the price -> ILUC land-use model (full precision, no rounding).
handoff_2035 <- price_shock |> dplyr::filter(year == 2035)

dir.create(file.path("data", "derived"), showWarnings = FALSE, recursive = TRUE)
utils::write.csv(
  handoff_2035,
  file.path("data", "derived", "vegetable_oil_price_shock_2035.csv"),
  row.names = FALSE
)
utils::write.csv(
  price_shock,
  file.path("data", "derived", "vegetable_oil_price_shock.csv"),
  row.names = FALSE
)

cat(sprintf(
  paste0(
    "2035 hand-off: VO demand %.0f billion L on a Q0 of %.0f billion L ",
    "(demand share %.0f%%) x multiplier %.2f = central price shock of +%.0f%%."
  ),
  handoff_2035$vo_demand_billion_liters,
  q0_billion_liters,
  handoff_2035$demand_share_pct,
  m_chen,
  handoff_2035$pct_delta_p_central
))
```

The 2035 demand share is far beyond the marginal regime, so the central shock should be read as "the price response is very large — a multiple of baseline prices" rather than as a point estimate; the land-use model runs will carry this caveat with them. The exported `pct_delta_p_central` column in `data/derived/vegetable_oil_price_shock_2035.csv` is the input to the next step of the analysis: converting the price shock into ILUC area and emissions.

## Discussion

*Discussion of findings will go here.*

## References

::: {#refs}
:::
