---
title: "Assessing the ILUC implications of increasing marine biofuel demand"
subtitle: "Prototype Proof-of-Concept Analyis"
authors: Environmental Markets Lab (emLab) at University of California, Santa Barbara (UCSB). For International Council on Clean Transportation (ICCT)
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
    embed-resources: false
bibliography: references.bib
---

## Overview

This report documents a protype proof-of-concept analysis investigating the projected indirect land use change (ILUC) implications of increasing marine biofuel demand. This prototype is being done by Environmental Markets Lab (emLab) at University of California, Santa Barbara (UCSB) for International Council on Clean Transportation (ICCT). All results presented in this analysis should be considered exploratory and preliminary.

## Background

Indirect land use change (ILUC) is a key factor in evaluating the lifecycle greenhouse gas emissions of biofuels. This prototype explores ILUC dynamics in the context of marine biofuel production.

## Methods

*Methods will be described here.*

## Data sources

### Projected demand for soy-based vegetable oil (from ICCT)

In the research brief
[Estimating the maximum potential vegetable oil demand for international shipping under the International Maritime Organization’s Net-Zero Framework](https://theicct.org/wp-content/uploads/2025/09/ID-450-Vegetable-Oil_research-brief_final.pdf) [@seber2025], ICCT estimates the projected demand for marine renewable biodiesel (and the associated demand in vegatable oil) under the proposed IMO Net Zero Framework, from 2018 to 2035. This project demand is summarized in the report's **Table 1** (page 5). We scraped the PDF report and turned this into a data tibble for use in our analysis:
```{r}
#| label: scrape-icct-table-1
#| message: false
#| warning: false

# Package versions are pinned via renv (see renv.lock); run renv::restore() to install.
# Functions are called with explicit pkg:: prefixes rather than attaching via library().

pdf_path <- file.path(
  "data",
  "icct_biofuel_demand_report",
  "ID-450-Vegetable-Oil_research-brief_final.pdf"
)

# Extract page 5 as text and keep only the data rows (years 2028–2035)
page5 <- pdftools::pdf_text(pdf_path)[5]
lines <- stringr::str_trim(stringr::str_split(page5, "\n")[[1]])
data_lines <- stringr::str_subset(lines, "^20(2[89]|3[0-5])\\b")

# Each data row yields exactly 10 numeric tokens:
#   year, shipping demand, base target, fossil LNG %,
#   UCO-RD %, UCO-RD bl, VO-RD %, VO-RD bl, UCO demand bl, VO demand bl
# Each year appears on its own line in the extracted text, so we
#pull the numeric tokens per row and split the two compound columns
# ("% energy demand / billion liters") into separate fields.
nums <- stringr::str_extract_all(data_lines, "[0-9]+\\.?[0-9]*")
mat <- do.call(rbind, lapply(nums, function(x) as.numeric(x[seq_len(10)])))

icct_table_1 <- tibble::as_tibble(mat, .name_repair = "minimal")
names(icct_table_1) <- c(
  "year",
  "shipping_energy_demand_ej",
  "base_target_trajectory_gco2e_mj",
  "fossil_lng_pct",
  "ucord_pct_energy_demand",
  "ucord_billion_liters",
  "vord_pct_energy_demand",
  "vord_billion_liters",
  "uco_demand_billion_liters",
  "vo_demand_billion_liters"
)

icct_table_1 <- icct_table_1 |>
  dplyr::mutate(year = as.integer(year)) |>
  dplyr::arrange(year)
```

```{r}
#| label: icct-table-1-interactive
#| message: false
#| warning: false

DT::datatable(
  icct_table_1,
  rownames = FALSE,
  filter = "top",
  caption = 'Caption from the ICCT report: "UCO-RD and VO-RD blend ratios and volumes to comply with the base target emissions trajectory. Notes: UCO-based renewable diesel availability is used as a constraint, and the rest of the demand is assumed to come from soy oil-based renewable diesel. UCO and VO demand to reach the estimated renewable diesel volumes are included. Estimations for shipping energy demand for ships of ≥ 5,000 gross tonnage are from the ICCT’s POLARIS Mode"',
  options = list(pageLength = 10, scrollX = TRUE)
)
```

The most important metric for our analysis is the projected demand for soy-based vegetable oil (to be used renewable biodiesel), in billion liters, across the 2028–2035 horizon. It is this projected demand of soy-based vegetable oil that we believe will lead to a direct price shock that can be used in our land use model. Here we look at a time series of this projected demand:

```{r}
#| label: icct-demand-timeseries
#| message: false
#| warning: false

ts_data <- icct_table_1 |>
  dplyr::select(
    year,
    vo_demand_billion_liters
  )

dygraphs::dygraph(
  ts_data,
  main = "Projected demand for soy-based vegetable oil",
  ylab = "Billion liters"
) |>
  dygraphs::dySeries("vo_demand_billion_liters", label = "VO demand") |>
  dygraphs::dyOptions(strokeWidth = 2, drawPoints = TRUE, pointSize = 3) |>
  dygraphs::dyLegend(show = "always") |>
  dygraphs::dyHighlight(highlightSeriesOpts = list(strokeWidth = 3))
```

## Results

*Results will go here.*

## Discussion

*Discussion of findings will go here.*

## References

::: {#refs}
:::
